---
title: "Practical Machine Learning Week 4"
author: "Harsh Gupta"
date: "10/22/2020"
output: html_document
---

## Summary

This report uses machine learning algorithms to predict the manner in which users of exercise devices exercise. 


### Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website [here:](http://groupware.les.inf.puc-rio.br/har) (see the section on the Weight Lifting Exercise Dataset). 

### Data 

The training data for this project are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv


# Loading the required libraries

```{r}
library(caret)
library(dplyr)
library(knitr)
library(VIM)
library(gbm)
library(ggplot2)
library(corrplot)

```


# Getting the data

```{r}
data_dir = "./data"
training_url = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test_url = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
training_file = "pml-training.csv"
test_file = "pml-test.csv"
if (!file.exists(data_dir)) {
  dir.create(data_dir)
}
if (!file.exists(file.path(data_dir, training_file))) {
  download.file(training_url, destfile=file.path(data_dir, training_file))
}
if (!file.exists(file.path(data_dir, test_file))) {
  download.file(test_url, destfile=file.path(data_dir, test_file))
}
```

## Reading the Data

Split the data into two dataframes. One for testing and one for training.

```{r}
train_data <- read.csv(file.path(data_dir, training_file))
test_data <- read.csv(file.path(data_dir, test_file))

```

## Cleaning and preprocessing the raw data

The dataset should be checked for any anomalies/ missing data so that our ML model is more accurate

```{r}
sum(complete.cases(train_data))
```



### Removing the columns with missing or NA values.


```{r}
colnames(train_data)
plot(colMeans(is.na(train_data)))
```

As we can see there are a number of columns with missing values.

Only the columns without any missing value/NA value is kept.


```{r}
trainClasse = train_data$classe
trainRaw = train_data[, sapply(train_data, is.numeric)]
testRaw = test_data[, sapply(test_data, is.numeric)]
```

Removing the columns with NA.

```{r}
trainFilter <- trainRaw[, colSums(is.na(trainRaw)) == 0]
# Attach Classe variable
trainFilter$classe = trainClasse
testFilter <- testRaw[, colSums(is.na(testRaw)) == 0]
```



Removing other useless columns.

```{r}
unwanted = !grepl("X|timestamp", colnames(trainFilter))
cols = colnames(trainFilter)[unwanted]
trainFilter = trainFilter %>%
  select(cols)
unwanted = !grepl("X|timestamp", colnames(testFilter))
cols = colnames(testFilter)[unwanted]
testFilter = testFilter %>%
  select(cols)


dim(trainFilter)
dim(testFilter)
```

## Splitting the training data into training and validation. 

We will split the Training data into Training and Validation set using the 80-20 rule. Validation data will be used to check the performance of model while training. And the final testing will be done on testing data.

```{r}
inTrain <- createDataPartition(trainFilter$classe, p=0.70, list=F)
trainData <- trainFilter[inTrain, ]
validationData <- trainFilter[-inTrain, ]
dim(trainData)
```

# Data modeling

We will be using  **Random Forest** and **XGBoost** ,which quite popular for initial training.

## Random forest Model


```{r}
set.seed(10000)
controlRf <- trainControl(method="cv", 5, allowParallel = TRUE)
modelRf <- train(classe ~ ., data=trainData, method="rf", ntree=50)
modelRf
```

### Testing done on validation set

```{r}
predict_rf <- predict(modelRf, validationData)
confusionMatrix(validationData$classe, predict_rf)
```

As the results show,it classified **classe** feature quite accurately.

## XGBoost

```{r}
controlXGB <- trainControl(method="cv", 5, allowParallel = TRUE)
modelXGB <- train(classe ~ ., data=trainData, method="xgbTree", trControl=controlXGB)
```

```{r}
modelXGB
```

### Testing done on validation set

```{r}
predict_XGB <- predict(modelXGB, validationData)
confusionMatrix(validationData$classe, predict_XGB)
```

XGB gave us even better result with only couple of misclassification

# Comparing the two  models

```{r}
# collect resamples
model_results <- resamples(list(RF=modelRf, XGB=modelXGB))
# summarize the distributions
summary(model_results)
# boxplots of results
bwplot(model_results)
# dot plots of results
dotplot(model_results)
```

# Final testing on testing dataset

```{r}
resultRf <- predict(modelRf, testFilter[, -length(names(testFilter))])
resultXGB <- predict(modelXGB, testFilter[, -length(names(testFilter))])
resultRf
resultXGB
confusionMatrix(resultRf, resultXGB)
```

# Conclusion

After testing the models on testing dataset we noticed that XGB works better with the trainig set.
